---
title: "stor538_proj2"
author: "Casey Dulaney"
date: "2023-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(rvest)
library(lubridate)
library(hms)
library(nbastatR)
library(xgboost)
library(purrr)
library(Metrics)
```


## Read in Files
```{r}
games <- read_csv(url("https://github.com/SuperMarioGiacomazzo/STOR538_WEBSITE/raw/master/Playoffs/Round%202/Starting%20Data/games.csv"))
games_details <- read_csv(url("https://github.com/SuperMarioGiacomazzo/STOR538_WEBSITE/raw/master/Playoffs/Round%202/Starting%20Data/games_details.csv"))
players <- read_csv(url("https://github.com/SuperMarioGiacomazzo/STOR538_WEBSITE/raw/master/Playoffs/Round%202/Starting%20Data/players.csv"))
ranking <- read_csv(url("https://github.com/SuperMarioGiacomazzo/STOR538_WEBSITE/raw/master/Playoffs/Round%202/Starting%20Data/ranking.csv"))
teams <- read_csv(url("https://github.com/SuperMarioGiacomazzo/STOR538_WEBSITE/raw/master/Playoffs/Round%202/Starting%20Data/teams.csv"))

```
















# Fixing Variable Type
```{r}


games =
games %>%
 mutate(
  spread = PTS_home - PTS_away,
  total = PTS_home + PTS_away,
  OREB = REB_home + REB_away)


games.grouped =
  games_details %>%
  group_by(GAME_ID, TEAM_ID) # grouping data by Game ID and Team ID so that we can analyze each separate team



# changing NAs in numeric variables to 0
games.grouped$FGM[is.na(games.grouped$FGM)] = 0
games.grouped$FGA[is.na(games.grouped$FGA)] = 0
games.grouped$FG_PCT[is.na(games.grouped$FG_PCT)] = 0
games.grouped$FG3M[is.na(games.grouped$FG3M)] = 0
games.grouped$FG3A[is.na(games.grouped$FG3A)] = 0
games.grouped$FG3_PCT[is.na(games.grouped$FG3_PCT)] = 0
games.grouped$FTM[is.na(games.grouped$FTM)] = 0
games.grouped$FTA[is.na(games.grouped$FTA)] = 0
games.grouped$FT_PCT[is.na(games.grouped$FT_PCT)] = 0
games.grouped$OREB[is.na(games.grouped$OREB)] = 0
games.grouped$DREB[is.na(games.grouped$DREB)] = 0
games.grouped$REB[is.na(games.grouped$REB)] = 0
games.grouped$AST[is.na(games.grouped$AST)] = 0
games.grouped$STL[is.na(games.grouped$STL)] = 0
games.grouped$BLK[is.na(games.grouped$BLK)] = 0
games.grouped$TO[is.na(games.grouped$TO)] = 0
games.grouped$PF[is.na(games.grouped$PF)] = 0
games.grouped$PTS[is.na(games.grouped$PTS)] = 0
games.grouped$PLUS_MINUS[is.na(games.grouped$PLUS_MINUS)] = 0

```


# fixing MIN variable
```{r}
games.grouped$MIN <- gsub(":", ".", games.grouped$MIN) # takes out colon of the Minutes variable; important for cleaning

games.grouped <- games.grouped %>%           # Create duplicate of column with new name
  mutate(SEC = MIN) 


# if the length of digits in MIN == 1, add a leading zero



# creating new "Minutes Played" variable that lets us express the minutes played variable as a numerical quantity
games.grouped$MIN = gsub('.{3}$', '', games.grouped$MIN)
games.grouped$SEC <- gsub("^.{0,2}", "", games.grouped$SEC)

games.grouped$MIN[is.na(games.grouped$MIN)] <- 0
games.grouped$SEC[is.na(games.grouped$SEC)] <- 0

games.grouped$MIN = as.numeric(games.grouped$MIN)
games.grouped$SEC = as.numeric(games.grouped$SEC)
x <- games.grouped$MIN
games.grouped$MIN = 
  ifelse (nchar(x)==1, paste0("0", x), x)

games.grouped$SEC = ((games.grouped$SEC) / 60) * 100
games.grouped$SEC <- gsub("^.{0,1}", "", games.grouped$SEC)
games.grouped$MinutesPlayed = paste0(games.grouped$MIN, games.grouped$SEC)

head(games.grouped)





```





# finalizing games grouped
```{r}

games.grouped = games.grouped  %>%
    group_by(GAME_ID, TEAM_ID) %>%
    mutate(rank = rank(desc(MinutesPlayed)))
  


  
head(games.grouped)

```




# Splitting Rebounds homes and away
```{r}

grouped.oreb <- summarize(games.grouped, sum(OREB))
grouped.dreb <- summarize(games.grouped, sum(DREB), sum(BLK), sum(TO), sum(PTS), sum(PF), sum(PTS), sum(AST), sum(REB), sum(FTA), sum(FTM), sum(FG3A), sum(FG3M), sum(FGM), sum(FGA))
grouped.oreb = grouped.oreb %>% 
  mutate(merging.id = paste(GAME_ID,TEAM_ID))

grouped.dreb = grouped.dreb %>% 
  mutate(merging.id = paste(GAME_ID,TEAM_ID))

```
  
  
# splitting games dataframe into home and away
```{r}
games.away = games
games.home = games

games.away = 
  games.away %>%
  select(GAME_DATE_EST, GAME_ID, TEAM_ID = VISITOR_TEAM_ID, SEASON, PTS = PTS_away, FG_PCT = FG_PCT_away, FG3_PCT = FG3_PCT_away, AST = AST_away, TOTAL_REB = REB_away)

games.home =
  games.home %>%
  select(GAME_DATE_EST, GAME_ID, TEAM_ID = HOME_TEAM_ID, SEASON, PTS = PTS_home, FG_PCT = FG_PCT_home, FG3_PCT = FG3_PCT_home, AST = AST_home, TOTAL_REB = REB_home)

games.all = rbind(games.home,games.away)
```


# merging rebounds into games
```{r}

games.all = games.all %>%
  mutate(merging.id=paste(GAME_ID, TEAM_ID))

games.all = na.omit(games.all)
games.all = games.all[!duplicated(games.all$merging.id), ]


all.rebounds = merge(grouped.oreb, grouped.dreb, by="merging.id")
all.rebounds = na.omit(all.rebounds)
all.rebounds = all.rebounds[!duplicated(all.rebounds$merging.id), ]


#games.rebounds.added <- merge(games.all, grouped.oreb, by="merging.id")
#games.rebounds.added <- merge(games.rebounds.added, grouped.dreb, by="merging.id")
#head(games.rebounds.added)
  
```



# final data frame: split by team and game
```{r}
games.split = merge(games.all, all.rebounds, by="merging.id")


games.split = games.split %>%
  transmute(merging.id = merging.id, DATE = GAME_DATE_EST, GAME_ID = GAME_ID.x, TEAM_ID = TEAM_ID.x, OREB =`sum(OREB)`, DREB = `sum(DREB)`, BLK = `sum(BLK)`, TO = `sum(TO)`, PTS = `sum(PTS)`, PF = `sum(PF)`, AST = `sum(AST)`, TOTAL_REB = `sum(REB)`, FTA = `sum(FTA)`, FTM = `sum(FTM)`, FG3M = `sum(FG3M)`, FG3A = `sum(FG3A)`, FGM = `sum(FGM)`, FGA = `sum(FGA)`, FG_PCT = FG_PCT, FG3_PCT = FG3_PCT)

# NOTE: this data frame has each row as a separate grouping (each row is a separate team that played in the game. For example, if the Hornets played the Kings, both teams are their own row)
```


## merging games split with team name
```{r}
games.split = merge(games.split, teams, by="TEAM_ID")

games.split = games.split %>%
  transmute(merging.id, DATE, GAME_ID, TEAM_ID, OREB, DREB, BLK, TO, PTS, PF, AST, TOTAL_REB, FTA, FTM, FG3M, FG3A, FGM, FGA, FG_PCT, FG3_PCT, TEAM_NAME = paste0(CITY," ", NICKNAME))
```



# final data frame: split by only game
```{r}
total.per.game = games.split %>%
 group_by(GAME_ID) %>%
summarize(sum(OREB), sum(DREB), sum(BLK), sum(TO), sum(PTS), sum(PF), sum(AST), sum(TOTAL_REB), sum(FTA), sum(FTM), sum(FG3M), sum(FG3A), sum(FGM), sum(FGA)) %>%
transmute(GAME_ID, OREB =`sum(OREB)`, DREB = `sum(DREB)`, BLK = `sum(BLK)`, TO = `sum(TO)`, PTS = `sum(PTS)`, PF = `sum(PF)`, AST = `sum(AST)`, TOTAL_REB = `sum(TOTAL_REB)`, FTA = `sum(FTA)`, FTM = `sum(FTM)`, FG3M = `sum(FG3M)`, FG3A = `sum(FG3A)`, FGM = `sum(FGM)`, FGA = `sum(FGA)`)
```



# Outside Source and New Variable
```{r}
gamedata <- game_logs(seasons = 2023,
                      league = "NBA",
                      result_types = "team",
                      season_types = "Regular Season",
                      nest_data = F,
                      assign_to_environment = TRUE,
                      return_message = TRUE)

gamedata_final = gamedata[,!names(gamedata) %in% c("urlTeamSeasonLogo",
                                                   "slugSeason",
                                                   "slugLeague",
                                                   "yearSeason",
                                                   "typeSeason")]
head(gamedata_final)
nrow(gamedata_final)
ncol(gamedata_final)
```






# Merging Outside Source and New Variable
```{r}
# changing variable names
gamedata_final2 = gamedata_final %>%
  mutate(merging.id = paste(idGame,idTeam)) %>%
  transmute(merging.id = merging.id, FGM = fgmTeam, FGA = fgaTeam, FG3M = fg3mTeam, FG3A = fg3aTeam, FTM = ftmTeam, FTA = ftaTeam, OREB = orebTeam, DREB = drebTeam, TOTAL_REB = trebTeam, AST = astTeam, BLK = blkTeam, PF = pfTeam, PTS = ptsTeam, TO = tovTeam, FG_PCT = pctFG2Team, FG3_PCT = pctFG3Team, DATE = dateGame, GAME_ID = idGame, TEAM_ID = idTeam, TEAM_NAME = nameTeam) 
```



# Total Game Day Data
```{r}
total_gamedata <- rbind(games.split, gamedata_final2)



```


# adding new variable
```{r}
total_gamedata$oPPFGteam = (total_gamedata$FGM * 2 + total_gamedata$FG3M * 3) / (total_gamedata$FGA)
total_gamedata$oPPAteam = (total_gamedata$FTM * 1 + total_gamedata$FGM * 2 + total_gamedata$FG3M * 3) / (total_gamedata$FGA + total_gamedata$FTA)
head(total_gamedata)
```


```{r}
total_gamedata = total_gamedata %>% group_by(GAME_ID)

total_gamedata = total_gamedata[!duplicated(total_gamedata$merging.id), ]

total_gamedata2 = total_gamedata %>% 
  summarize(TOTAL_PTS = sum(PTS), SPREAD = last(PTS) - first(PTS), TOTAL_OREB = sum(OREB))

total_gamedata = merge(total_gamedata, total_gamedata2, by="GAME_ID")

```





________________________________________

# Modelling


## testing and training
```{r}
require(xgboost)
set.seed(538)
sample <- sample(c(TRUE, FALSE), nrow(total_gamedata), replace=TRUE, prob=c(0.7,0.3))
train  <- total_gamedata[sample, ]
test   <- total_gamedata[!sample, ]
glimpse(train)
```



### Modelling attempt 2
```{r}
library(caTools)
set.seed(538)
sample_split <- sample.split(Y = total_gamedata$PTS, SplitRatio = 0.7)
train_set <- subset(x = total_gamedata, sample_split == TRUE)
test_set <- subset(x = total_gamedata, sample_split == FALSE)

y_train <- as.integer(total_gamedata$PTS) - 1
y_test <- as.integer(total_gamedata$PTS) - 1
X_train <- total_gamedata %>% select(OREB,DREB,BLK,TO,PF,AST,FTA,FTM,FG3M,FG3A,FGM,FGA,FG_PCT,FG3_PCT,oPPFGteam,oPPAteam)
X_test <- total_gamedata %>% select(OREB,DREB,BLK,TO,PF,AST,FTA,FTM,FG3M,FG3A,FGM,FGA,FG_PCT,FG3_PCT,oPPFGteam,oPPAteam)


xgb_train <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
xgb_test <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)
#params <- list(
 #       booster = "gbtree",
 #       objective = "multi:softprob",
  #      eta=0.3,
  #      gamma=0,
   #     max_depth=6,
    #    min_child_weight=1,
     #   subsample=1,
      #  colsample_bytree=1
#)


#xgbcv <- xgb.cv(params = params
 #               ,data = xgb_train
  #              ,nrounds = 100
   #             ,nfold = 5
    #            ,showsd = T
    #            ,stratified = T
    ##            ,print.every.n = 10
     #           ,early.stop.round = 20
      #          ,maximize = F
#)
  
```

```{r}
watchlist = list(train=xgb_train, test=xgb_test)
xgb_model = xgb.train(data = xgb_train, max.depth = 3, watchlist=watchlist, nrounds = 70)
xgb_model





```


```{r}
library(caret)
final = xgboost(data = xgb_train, max.depth = 3, nrounds = 56, verbose = 0)
pred_y = predict(final, xgb_test)
final

caret::MAE(y_test, pred_y)
```

```{r}
importance_matrix <- xgb.importance(
  feature_names = colnames(xgb_train), 
  model = xgb_model
)
importance_matrix
```

```{r}
xgb.plot.importance(importance_matrix)
```
```{r}
xgb_preds <- predict(xgb_model, as.matrix(X_test), reshape = TRUE)
xgb_preds <- as.data.frame(xgb_preds)
colnames(xgb_preds) <- levels(total_gamedata$PTS)


xgb_preds$PredictedClass <- apply(xgb_preds, 1, function(y) colnames(xgb_preds)[which.max(y)])
xgb_preds$ActualClass <- levels(total_gamedata$PTS)[y_test + 1]



accuracy <- sum(xgb_preds$PredictedClass == xgb_preds$ActualClass) / nrow(xgb_preds)
accuracy
```

```{r}
pred <- predict(xgb_model, xgb_test)
```


_______
# Linear Model
```{r}


gamedata_final3 = filter(total_gamedata, GAME_ID %in% (22200001:22201170))

data23 = gamedata_final3 %>% 
  group_by(TEAM_ID) %>%
  summarize(OREB=mean(OREB),DREB=mean(DREB),BLK=mean(BLK),TO= mean(TO),PTS=mean(PTS),PF= mean(PF), AST=mean(AST), TOTAL_REB=mean(TOTAL_REB), FTA=mean(FTA), FTM=mean(FTM), FG3M=mean(FG3M), FG3A=mean(FG3A), FGM=mean(FGM), FGA=mean(FGA), FG_PCT=mean(FG_PCT), FG3_PCT=mean(FG3_PCT), oPPFGteam=mean(oPPFGteam), oPPAteam=mean(oPPAteam), TOTAL_PTS=mean(TOTAL_PTS))

```


```{r}

linearmodel <- lm(PTS~OREB+DREB+BLK+TO+PF+AST+TOTAL_REB+FTA+FTM+FG3M+FG3A+FGM+FGA+FG_PCT+FG3_PCT+oPPFGteam+oPPAteam, data=total_gamedata)
summary(linearmodel)

predictions_PTS_linear = predict(linearmodel,data23)
predictions_PTS_linear
```
```{r}
xgb_preds2 <- predict(xgb_model, as.matrix(X_test), reshape = TRUE)
xgb_preds2 <- as.data.frame(xgb_preds)
xgb_preds2
```



# Linear Training
```{r}
mod2 <- lm(PTS~oPPAteam+OREB+DREB+BLK+PF+AST+FTA+FG3A+FG_PCT+FG3_PCT+oPPFGteam+oPPAteam, data=total_gamedata)
#mod3 <- glm(PTS~ data=total_gamedata, family="poisson")
summary(mod3)
mae(total_gamedata$PTS, predict(linearmodel))
mae(total_gamedata$PTS, predict(mod2))
mae(total_gamedata$PTS, predict(mod3))

```

```{r}
predict(linearmodel,data23)
predict(mod3,data23)
predict(mod2,data23)
```

```{r}
require(xgboost)
sample <- sample(c(TRUE, FALSE), nrow(total_gamedata), replace=TRUE, prob=c(0.7,0.3))
train  <- total_gamedata[sample, ]
test   <- total_gamedata[!sample, ]
#sparse_matrix <- sparse.model.matrix(response ~ .-1, data = train)
#xgtrain <- xgb.DMatrix(data = as.matrix(train[,-5]), label = train$OREB)
train2 <- select(train, -c(DATE, TEAM_NAME, merging.id))
dtrain = xgb.DMatrix(as.matrix(train2), label=as.matrix(train2$PTS))
```

```{r}
y_train <- train2$PTS
y_test <- test$PTS
lm_train <- select(train2, -c(PTS, FGM, FG3M, FTM))
lm_test <- select(test, -c(PTS, FGM, FG3M, FTM))
model = lm(y_train~DREB+BLK+TO+PF+AST+FTA+OREB+FGA+FG3_PCT+oPPAteam, data=lm_train)
summary(model)

```

```{r}

lm_total_predict_train = predict(model, lm_train) ##make predictions on data
lm_total_predict_test = predict(model, lm_test)
Metrics::mae(lm_total_predict_train, y_train) #evaluate your predictions based on MAE
Metrics::mae(lm_total_predict_test, y_test)
```

#Stepwise Regression
```{r}
train3 = select(train2, -c(GAME_ID, TOTAL_REB, DREB, TEAM_ID, FGM, FTM, FG3M, TOTAL_POINTS))

full_model = lm(PTS~PF+TO+AST+FTA+FG3A+FGA+FG3_PCT+oPPAteam+FG3A*FGA, data = train3)
# Stepwise regression model
step_model = MASS::stepAIC(full_model, direction = "both",trace = F)
summary(step_model)
```

```{r}

lm_total_predict_train = predict(step_model, lm_train) ##make predictions on data
lm_total_predict_test = predict(step_model, lm_test)
Metrics::mae(lm_total_predict_train, y_train) #evaluate your predictions based on MAE
Metrics::mae(lm_total_predict_test, y_test)
```
# plot the best fit model




#RandomForest
```{r}
library(randomForest)

set.seed(1)

forest_model = randomForest(formula=PTS~., data=train3, metric='mae',ntree=4, importance=T)
importance(forest_model)[order(-importance(forest_model)[, 1]),]
varImpPlot(forest_model)
```

```{r}

rf_predict_train = predict(forest_model, train)
rf_predict_test = predict(forest_model, test)

Metrics::mae(rf_predict_train, train$PTS)
Metrics::mae(rf_predict_test, test$PTS)

```
#Polynomial
```{r}

model2 = lm(PTS~poly(FGA,3)+poly(oPPAteam,2)+poly(oPPFGteam,2)+AST+TO+FG_PCT, data=train3)
summary(model2)
rf_predict_train = predict(model2, train)
rf_predict_test = predict(model2, test)

Metrics::mae(rf_predict_train, train$PTS)
Metrics::mae(rf_predict_test, test$PTS)
```
# using data23 to predict with our model
```{r}
predict(step_model,data23)
TOTAL_PTS_PREDICTIONS$predict = predict(step_model, newdata=data23)

```


```{r}
data23
```


```{r}

df_pts <- data.frame(first_column=predict(step_model, newdata=data23), second_column=
  c("Atlanta Hawks","Boston", "Pelicans","Bulls", "Mavericks", "Nuggets", "Rockets","Clippers","Lakers", "Heat", "Bucks", "Timberwolves","Nets","Knicks", "Magic", "Pacers", "76ers", "Suns", "Blazers", "Kings", "Spurs","Thunder", "Raptors", "Jazz", "Grizzlies","Wiz","Pistons","Hornets","Cavs", "Warriors"))
```







```


